---
title: "Additional Exploration for Prescribers"
output: html_notebook
---

### Tennessee's Opioid Crisis

Opioids are a class of drugs including prescription pain relievers such as oxycodone and hydrocodone, the synthetic opioid fentanyl, and the illegal drug heroin. These drugs produce euphoria in addition to pain relief, which can lead to dependence, addiction, overdose, and death. 

In the late 1990s, opioids began to be prescribed at high rates, which led to widespread misuse and ultimately created a serious national health crisis. In 2019, more than 130 people per day died from opioid-related drug overdoses in the United States. Tennessee has been one of the hardest-hit states in the opioid epidemic, with more than 1300 overdose deaths in 2018.

In this project, you will be working with a database created from the 2017 Medicare Part D Prescriber Public Use File, available at https://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/Medicare-Provider-Charge-Data/PartD2017. 

Your objectives are to use the database provided to attempt to answer the following questions:

```{r, echo = FALSE, warrnings = FALSE}
library(DBI)
library(RPostgres)
library(tidyverse)
library(knitr)

con <- dbConnect(Postgres(),                 
                 dbname = 'prescribers',
                 host = 'localhost',
                 port = 5432, 
                 user = 'postgres',
                 password = rstudioapi::askForPassword("Database Password: "))
```

* Which Tennessee counties had a disproportionately high number of opioid prescriptions? **Scott, Clay, Washington, etc.** 

```{r, echo = FALSE, warrnings = FALSE}
query = "
WITH correct_counties AS (
	SELECT 
		zip,
		fipscounty
	FROM zip_fips
	WHERE tot_ratio IN (
			SELECT MAX(tot_ratio) OVER(PARTITION BY zip) AS max_tot
			FROM zip_fips)
),
applicable_npis AS (
	SELECT npi, county, population
	FROM prescriber
	INNER JOIN correct_counties
	ON nppes_provider_zip5 = zip
	INNER JOIN population
	USING (fipscounty)
	INNER JOIN fips_county
	USING(fipscounty)
	WHERE nppes_provider_state = 'TN'
)

SELECT county, ROUND(SUM(total_claim_count)/population, 2) AS proportion_opioid_prescriptions
FROM drug
RIGHT JOIN prescription
USING (drug_name)
INNER JOIN applicable_npis
USING (npi)
WHERE opioid_drug_flag = 'Y'
GROUP BY county, population
ORDER BY proportion_opioid_prescriptions DESC;
"
res <- dbSendQuery(con, query)
proportion_opioid_prescriptions <- dbFetch(res)

proportion_opioid_prescriptions %>% 
  slice(1:10) %>% 
  ggplot(aes(reorder(str_to_title(county), 
                     proportion_opioid_prescriptions), 
             proportion_opioid_prescriptions,
             fill = county)) +
           geom_bar(stat = 'identity') +
  coord_flip() +
  labs(title = "Top 10 TN Counties by Proportion of Opioid Prescriptions",
       x = "Proportion of Opioid Prescriptions by Population",
       y = "County") +
  theme(legend.position = "None")
```
* Who are the top opioid prescribers for the state of Tennessee?
**Coffey, Kindrick, Cathers, Painter, Clark.**

```{r, echo = FALSE, warrnings = FALSE}
query = "
SELECT 
	INITCAP(nppes_provider_first_name) AS first_name,
	COALESCE(nppes_provider_mi,'') AS mi,
	INITCAP(nppes_provider_last_org_name) AS last_name, 
	COALESCE(nppes_credentials,'') AS credentials,
	SUM(total_claim_count) AS total_prescriptions
FROM drug
INNER JOIN prescription
USING (drug_name)
INNER JOIN prescriber
USING (npi)
WHERE opioid_drug_flag = 'Y'
AND nppes_provider_state = 'TN'
GROUP BY
	nppes_provider_first_name,
	nppes_provider_mi,
	nppes_provider_last_org_name, 
	nppes_credentials
ORDER BY total_prescriptions DESC;
"
res = dbSendQuery(con, query)
prescriptions_by_tn_providers <- dbFetch(res) %>% 
  as_tibble() %>%
  mutate(credentials = paste0(", ", str_remove_all(credentials, "\\.")),
         mi = if_else(mi == '', mi, paste0(mi,".")),
         provider_name = str_c(str_c(first_name, mi, last_name, sep = ' '), credentials)) %>%
  select(provider_name, total_prescriptions) %>% 
  mutate(provider_name = str_replace_all(provider_name, "  ", " "))


prescriptions_by_tn_providers %>% 
  slice(1:10) %>% 
  ggplot(aes(reorder(provider_name, 
                     total_prescriptions), 
             total_prescriptions,
             fill = provider_name)) +
           geom_bar(stat = 'identity') +
  coord_flip() +
  labs(title = "Top 10 Opioid Prescribers in TN",
       x = "Number of Opioid Prescriptions",
       y = "Prescriber") +
  theme(legend.position = "None")


```

* What did the trend in overdose deaths due to opioids look like in Tennessee from 2015 to 2018?
**Mild increase during that time from 1,000 to 1,300.**

```{r, echo = FALSE, warrnings = FALSE}
query = "
SELECT year, sum(overdose_deaths) AS deaths
FROM overdose_deaths
WHERE year BETWEEN 2015 AND 2018
GROUP BY year
ORDER BY year;
"

res <- dbSendQuery(con, query)
deaths <- dbFetch(res)

deaths %>% 
  ggplot(aes(year, deaths)) +
  geom_line() +
  ylim(0,1500) +
  labs(title = "Deaths Due to Opioid Overdoses in TN (2015-2018)",
       x = "Year",
       y = "Number of Deaths")

```

* Is there an association between rates of opioid prescriptions and overdose deaths by county? **Almost none. Correlation is nearly 0.**

```{r, echo = FALSE, warrnings = FALSE}
query = "
WITH correct_counties AS (
	SELECT 
		zip,
		fipscounty
	FROM zip_fips
	WHERE tot_ratio IN (
			SELECT MAX(tot_ratio) OVER(PARTITION BY zip) AS max_tot
			FROM zip_fips)
),
applicable_npis AS (
	SELECT npi, county, population
	FROM prescriber
	INNER JOIN correct_counties
	ON nppes_provider_zip5 = zip
	INNER JOIN population
	USING (fipscounty)
	INNER JOIN fips_county
	USING(fipscounty)
	WHERE nppes_provider_state = 'TN'
),
proportional_query AS (
	SELECT 
		county, 
		ROUND(SUM(total_claim_count)/population, 2) AS proportion_opioid_prescriptions
	FROM drug
	RIGHT JOIN prescription
	USING (drug_name)
	INNER JOIN applicable_npis
	USING (npi)
	WHERE opioid_drug_flag = 'Y'
	GROUP BY county, population
	ORDER BY proportion_opioid_prescriptions DESC
),
deaths_query AS (
	SELECT
		county,
		year,
		ROUND(overdose_deaths/population * 1000.0, 2) AS odds_per_1000
	FROM overdose_deaths
	INNER JOIN fips_county
	USING (fipscounty)
	INNER JOIN population
	USING (fipscounty)
	WHERE state = 'TN'
)

SELECT 
	county, 
	year, 
	odds_per_1000, 
	proportion_opioid_prescriptions
FROM proportional_query
INNER JOIN deaths_query
USING (county);
"
res <- dbSendQuery(con, query)
odds_vs_pns <- dbFetch(res)

odds_vs_pns %>% 
  filter(year == 2017) %>% 
  ggplot(aes(proportion_opioid_prescriptions, 
             odds_per_1000)) +
  geom_point(size = 2) +
  scale_x_continuous(limits = c(0,1.2),
                     breaks = seq(0,1.2,0.2)) +
  scale_y_continuous(limits = c(0,1.2),
                     breaks = seq(0,1.2,0.2)) +
  labs(x = "Opioid Prescriptions Per Capita",
       y = "Overdose Deaths Per 1,000",
       title = "Overdose Deaths vs. Opioid Prescriptions in TN")

# odds_vs_pns %>% 
#   filter(year == 2017) %>% 
#   select(proportion_opioid_prescriptions, odds_per_1000) %>% 
#   cor() %>% 
#   .[1,2]
```
* Is there any association between a particular type of opioid and number of overdose deaths?
**Not really. Hydrocodone and Oxycodone are the closest contenders, but they do not correlate with overdose deaths at the county level.**

```{r, echo = FALSE, warrnings = FALSE}
query = "
WITH deaths_query AS (
	SELECT
		county,
		year,
		ROUND(overdose_deaths/population * 1000.0, 2) AS odds_per_1000
	FROM overdose_deaths
	INNER JOIN fips_county
	USING (fipscounty)
	INNER JOIN population
	USING (fipscounty)
	WHERE state = 'TN'
),
renamed_drugs AS (
	SELECT 
		drug_name,
		generic_name,
		CASE WHEN generic_name ILIKE '%CODEINE%' THEN 'codeine'
		WHEN generic_name ILIKE '%FENTANYL%' THEN 'fentanyl'
		WHEN generic_name ILIKE '%HYDROCODONE%' THEN 'hydrocodone'
		WHEN generic_name ILIKE '%MORPHINE%' THEN 'morphine'
		WHEN generic_name ILIKE '%OXYCODONE%' THEN 'oxycodone'
		WHEN generic_name ILIKE '%OXYMORPHONE%' THEN 'oxymorphone'
		END AS generic_type
	FROM drug
),
drug_counts AS (
	SELECT
		npi,
		generic_type, 
		SUM(total_claim_count) AS drug_tcc
	FROM prescription
	INNER JOIN renamed_drugs
	USING (drug_name)
	WHERE generic_type IS NOT NULL
	GROUP BY npi, generic_type
),
correct_counties AS (
	SELECT 
		zip,
		fipscounty
	FROM zip_fips
	WHERE tot_ratio IN (
			SELECT MAX(tot_ratio) OVER(PARTITION BY zip) AS max_tot
			FROM zip_fips)
),
applicable_npis AS (
	SELECT npi, county, population
	FROM prescriber
	INNER JOIN correct_counties
	ON nppes_provider_zip5 = zip
	INNER JOIN population
	USING (fipscounty)
	INNER JOIN fips_county
	USING(fipscounty)
	WHERE nppes_provider_state = 'TN'
),
proportional_query AS (
	SELECT 
		generic_type,
		county, 
		ROUND(SUM(drug_tcc)/population, 2) AS proportion_prescriptions
	FROM drug_counts
	INNER JOIN applicable_npis
	USING (npi)
	GROUP BY county, population, generic_type
	ORDER BY county, generic_type
)

SELECT 
	INITCAP(county) AS county, 
	year, 
	generic_type,
	odds_per_1000, 
	proportion_prescriptions
FROM proportional_query
INNER JOIN deaths_query
USING (county)
ORDER BY county, generic_type, year;
"

res <- dbSendQuery(con, query)
odds_vs_types <- dbFetch(res)

odds_vs_types %>% 
  filter(year == 2017,
         generic_type == 'oxycodone' |
           generic_type == 'hydrocodone') %>% 
  ggplot(aes(proportion_prescriptions, 
             odds_per_1000,
             color = generic_type)) +
  geom_point(size = 2) + 
  scale_color_discrete(name = 'Type of Drug') +
  labs(Title = 'Correlation Between Hydrocodone and Oxycodone and Overdose Deaths',
       x = 'Scripts Per Capita',
       y = 'Overdose Deaths Per 1,000')

```

Note that some zip codes will be associated with multiple fipscounty values in the zip_fips table. To resolve this, use the fipscounty with the highest tot_ratio for each zipcode.

Use Python or R to do your analysis on these questions. 
